---
title: "Legislative Output of the European Union"
subtitle: "An Automatically Updated Overview of the EU's Legislative Activity"
author: Dr Michal Ovádek
output: 
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
---

<style>
body {
text-align: justify}
</style>

<style type="text/css">
  body{
  font-size: 13pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      error = FALSE,
                      fig.width = 9,
                      fig.height = 6)
```


```{r data, include=FALSE}

# packages
library(eurlex)
library(ggplot2)
library(dplyr)
library(stringr)
library(rmarkdown)
library(modelsummary)
library(ggiraph)

# seed
set.seed(65234)

# acts
acts_proposals <- elx_make_query("any", sector = 3,
                           include_date = TRUE,
                           include_proposal = TRUE) |>
  elx_run_query() |> 
  select(-work) |> 
  rename(date = `callret-3`)

# acts only
acts <- acts_proposals |> 
  filter(!is.na(celex)) |> 
  distinct(celex, .keep_all = T) |> 
  select(-proposal)

# 4 main types of acts
regs <- acts |> filter(str_sub(celex,6,6) == "R")
decs <- acts |> filter(str_sub(celex,6,6) == "D")
dirs <- acts |> filter(str_sub(celex,6,6) == "L")
recs <- acts |> filter(str_sub(celex,6,6) == "H")

# proposals
proposals <- elx_make_query("proposal", include_date = TRUE) |> 
  elx_run_query() |> 
  select(-work,-type) |> 
  rename(date_proposal = `callret-3`) |> 
  filter(!is.na(celex),
         !is.na(date_proposal)) |> 
  distinct(celex, .keep_all = TRUE)

# current date
date_now <- Sys.Date()
day_now <- as.integer(str_sub(date_now, 9, 10))
```

The European Union (EU) is often criticized for having a heavy regulatory touch and being unwieldy. What is the reality? This automatically updated research document gives a real-time overview of the EU's legislative output and efficiency.

As of ```r format(Sys.time(), '%d %B %Y')```, the EU and its predecessors have produced ```r nrow(regs)``` regulations, ```r nrow(dirs)``` directives, ```r nrow(decs)``` decisions and ```r nrow(recs)``` recommendations, according to [Eur-Lex](https://eur-lex.europa.eu/homepage.html) data pulled via the [eurlex](https://github.com/michalovadek/eurlex) package for R and a SPARQL API maintained by the Publications Office of the European Union.^[This also means that any omissions and mistakes present in Eur-Lex are carried through to the output shown here.]

## Number of acts over time {.tabset}

We start by looking at how the number of the four main legal acts changed over the lifespan of the EU. The number of acts adopted yearly is a parsimonious – but imperfect^[We should also take into account the regulatory breadth and depth of each legal act. A decision increasing the tariff on steel by 5 per cent is a very different act from, say, the General Data Protection Regulation. ] – proxy for the EU's regulatory tendencies.

### Plot

```{r ncounttime}

# select only main act types
relevant_acts <- bind_rows(Regulation = regs, Decison = decs, Directive = dirs, Recommendation = recs,
                        .id = "type") |> 
  mutate(year = as.integer(str_sub(celex, 2,5))) |> 
  mutate(month_year = as.Date(str_replace(date, "[:digit:][:digit:]$", "01")))

# count
n_relevant <- relevant_acts |> 
  count(year, type)

# viz
ggplot(n_relevant, aes(x = year, y = n, fill = type, color = type)) +
  geom_col(show.legend = FALSE) +
  scale_fill_brewer(palette = "Spectral") +
  scale_color_brewer(palette = "Spectral") +
  facet_wrap(~type, scale = "free_y") +
  theme_minimal() +
  theme(legend.position = "top",
        legend.justification = "left",
        legend.title = element_text(face = "italic"),
        plot.background = element_rect(fill = "white", color = "grey88"),
        panel.grid = element_line(color = "grey95"),
        axis.text = element_text(color = "grey10"),
        title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "italic"),
        strip.text = element_text(hjust = 0, face = "bold")) +
  labs(x = NULL,
       y = "Number of acts",
       title = "Number of legal acts produced by the European Union",
       subtitle = "Aggregated at the year level by date of publication")

# vars for text
year1 <- as.integer(str_sub(date_now,1,4))-1
year2 <- as.integer(str_sub(date_now,1,4))-2

n_year1 <- n_relevant |> 
  filter(year == year1) |> 
  summarise(n = sum(n, na.rm = T)) |> 
  as.integer()

n_year2 <- n_relevant |> 
  filter(year == year2) |> 
  summarise(n = sum(n, na.rm = T)) |> 
  as.integer()

nyeardif <- ifelse(n_year1 < n_year2, "less than", "more than")
nyeardif <- ifelse(n_year1 == n_year2, "the same as", nyeardif)

```

### Code

```{r ncounttime, eval=FALSE, echo=TRUE}
```

### Dataframe
```{r ncounttimedf}
n_relevant
```

## {-}

In ```r year1```, the EU produced ```r n_year1``` legal acts which is ```r nyeardif``` the ```r n_year2``` acts in ```r year2```. Overall, the EU's legislative output is currently considerably lower compared to, in particular, the 1980s and 1990s.

### Proportion of act types {.tabset}

So we see there is significant over-time variation in the use of the different types of legal acts, but some types of acts, notably regulations, have been historically significantly more common than other types of acts. We can visualize the evolution of proportions directly.

#### Plot

```{r proptime}

# create df
proptime_df <- expand.grid(year = min(n_relevant$year):max(n_relevant$year),
            type = unique(n_relevant$type)) |> 
  left_join(n_relevant) |> 
  mutate(n = ifelse(is.na(n), 0L, n)) |> 
  group_by(year) |> 
  mutate(total_y = sum(n),
         proportion = n/total_y) |> 
  ungroup()

# viz
proptime_df |> 
  ggplot(aes(x = year, y = proportion, fill = type, color = type)) +
  geom_area(show.legend = TRUE) +
  geom_vline(xintercept = c(1960,1980,2000,2020), lty = 2, color = "grey70") +
  scale_fill_brewer(palette = "Spectral") +
  scale_color_brewer(palette = "Spectral") +
  theme_minimal() +
  theme(legend.position = "top",
        legend.justification = "left",
        legend.title = element_text(face = "italic"),
        plot.background = element_rect(fill = "white", color = "grey88"),
        panel.grid = element_blank(),
        axis.text = element_text(color = "grey10"),
        title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "italic"),
        strip.text = element_text(hjust = 0, face = "bold")) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = NULL,
       y = NULL,
       color = NULL,
       fill = NULL,
       title = "Relative prevalence of legal acts",
       subtitle = "Proportion of each type of legal act in a given year",)

```

#### Code

```{r proptime, eval=FALSE, echo=TRUE}
```

#### Dataframe
```{r proptimedf}
proptime_df
```

### {-}

We can see that although regulations used to dominate the EU's legislative output, decisions – which are typically associated with various administrative actions – are nowadays almost equally prevalent.

The rise in the number of recommendations can be mostly attributed to the strengthening of the EU's role in the surveillance and coordination of the Member States' economic policies.

## Year-on-year comparison {.tabset}

The daily streaming of data can be relied on to make completely up-to-date comparisons of EU legislative activity. For example, we can look at the extent to which the monthly adoption rate of legal acts this year differs from last year.

### Plot

```{r yearonyear}

# monthly output last two years
last_two_years <- relevant_acts |> 
  mutate(year = as.integer(str_sub(date, 1,4))) |> 
  filter(year %in% c(max(relevant_acts$year), max(relevant_acts$year)-1)) |> 
  count(month_year, type)

last_two_years_full <- expand.grid(month_year = unique(last_two_years$month_year),
                                   type = unique(last_two_years$type)) |> 
  left_join(last_two_years) |> 
  mutate(n = ifelse(is.na(n), 0L, n))

# year on year change
last_two_yoy_diff <- last_two_years_full |> 
  arrange(month_year) |> 
  mutate(month = str_sub(month_year, 6,7)) |> 
  group_by(month, type) |> 
  summarise(yoy_diff = diff(n)) |> 
  ungroup()

# this month for text
this_month_yoy <- last_two_yoy_diff |> 
  arrange(desc(month)) |> 
  slice(1:4) |> 
  summarise(sum = sum(yoy_diff)) |> 
  as.integer()

this_month_yoy_dir <- ifelse(this_month_yoy < 0, "fewer than", "more than")
#this_month_yoy_dir <- ifelse(this_month_yoy == 0, "the same as", this_month_yoy)

this_month_yoy_abs <- abs(this_month_yoy)

# viz
last_two_yoy_diff |> 
  ggplot(aes(x = month, y = yoy_diff, fill = type, color = type)) +
  geom_col(show.legend = FALSE) +
  geom_hline(yintercept = 0, color = "grey83", lty = 1) +
  facet_wrap(~type, scales = "free_y") +
  scale_fill_brewer(palette = "Spectral") +
  scale_color_brewer(palette = "Spectral") +
  theme_minimal() +
  theme(legend.position = "top",
        legend.justification = "left",
        legend.title = element_text(face = "italic"),
        plot.background = element_rect(fill = "white", color = "grey88"),
        panel.grid = element_blank(),
        axis.text = element_text(color = "grey10"),
        title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "italic"),
        strip.text = element_text(hjust = 0, face = "bold")) +
  labs(x = "Month", y = "Year-on-year difference",
       title = "Monthly year-on-year comparison",
       subtitle = "Difference in the number of acts adopted this year compared to last year")

```

### Code

```{r yearonyear, eval=FALSE, echo=TRUE}
```

### Dataframe
```{r yoydf}
last_two_yoy_diff
```

## {-}

So far this month, the EU has adopted ```r this_month_yoy_abs``` legal acts ```r this_month_yoy_dir``` the same month last year. `r if(day_now < 16){"But we are still early in the month."}`

## Legislative efficiency {.tabset}

The image of the EU as an unwieldy bureaucracy with a complicated legislative system appears often in national media. Moreover, most theoretical literature on political systems would expect the number of veto players to affect legislative efficiency. On this basis, we formulate the simple hypothesis that 

> H1: The more EU Member States there are, the longer it takes to pass legislation.

Most people are therefore susceptible to assume that the EU needs a lot of time to adopt legal acts. Let's use Eur-Lex data to find out just how long it normally takes to pass legislation.

### Plot

```{r efficiencyviz}

# calculate maximum length of adoption and discard negative
acts_days <- acts_proposals |> 
  filter(!is.na(celex),
         !is.na(date),
         !is.na(proposal)) |> 
  left_join(proposals, by = c("proposal"="celex")) |> 
  mutate(days = as.integer(as.Date(date) - as.Date(date_proposal))) |> 
  filter(days > -1) |> 
  group_by(celex) |> 
  filter(days == max(days)) |> 
  ungroup() |> 
  distinct(celex, .keep_all = TRUE) |> 
  arrange(date) |> 
  mutate(year_adopted = as.integer(str_sub(date, 1, 4)),
         type = str_sub(celex, 6, 6),
         type = case_when(type == "R" ~ "Regulation",
                          type == "L" ~ "Directive",
                          type == "D" ~ "Decision",
                          T ~ "Other"),
         five_year = cut_interval(year_adopted, length = 5)) |> 
  filter(year_adopted > 1985) # not many complete pairs before this

# identify outliers
outliers <- acts_days |> 
  group_by(type) |> 
  mutate(max = max(days),
         mean = mean(days),
         coef = days / mean) |> 
  filter(coef > 2.9) |> 
  ungroup()

# plot
acts_days |> 
  filter(!celex %in% outliers$celex) |> 
  group_by(type) |> 
  mutate(global_mean = mean(days)) |> 
  ggplot(aes(x = five_year, y = days, color = type, fill = type)) +
  geom_boxplot(alpha = 0.1, show.legend = FALSE) +
  geom_hline(aes(yintercept = global_mean), lty = 2, color = "grey70") +
  facet_wrap(~type, dir = "h", ncol = 1, scales = "free_y") +
  scale_fill_brewer(palette = "Spectral") +
  scale_color_brewer(palette = "Spectral") +
  theme_minimal() +
  theme(legend.position = "top",
        legend.justification = "left",
        legend.title = element_text(face = "italic"),
        plot.background = element_rect(fill = "white", color = "grey88"),
        panel.grid = element_blank(),
        axis.text = element_text(color = "grey10"),
        title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "italic"),
        plot.caption = element_text(face = "italic", size = 8),
        strip.text = element_text(hjust = 0, face = "bold")) +
  labs(x = NULL,
       y = "Days to adoption",
       title = "Number of days between proposal and adoption",
       subtitle = "By five-year window in which the act was adopted*",
       caption = "* Outliers are pruned from the analysis, dashed line shows the global mean")

```

### Code

```{r efficiencyviz, echo=TRUE, eval=FALSE}
```

### Dataframe

```{r efficiencyvizdf}

acts_days

```

## {-}

Directives take by far the longest to adopt; unlike decisions and regulations, they are rarely deployed to regulate trivial matters. Nonetheless, there are few easily discernible temporal patterns in the plot. Most of all, it is not obvious that legislative efficiency would decrease as the EU expanded its membership.

However, before moving onto further investigation of the efficiency hypothesis, we should pause and take a look at some of the discarded outliers. What legislation took the longest to adopt?

```{r outliers, fig.width=10, fig.height=8}

# get titles for top outliers
outliers_top_txt <- outliers |> 
  arrange(-days) |> 
  slice(1:70) |> 
  rowwise() |> 
  mutate(title = elx_fetch_data(paste("http://publications.europa.eu/resource/celex/", celex, sep = ""),
                                "title")) |> 
  ungroup()

# interactive plot
iplot_outliers <- outliers_top_txt |> 
  bind_rows(data.frame(type = "Recommendation")) |> 
  ggplot(aes(x = as.Date(date), y = days, color = type, fill = type)) +
  geom_point_interactive(aes(tooltip = title, data_id = celex), 
                         alpha = 0.5,
                         size = 3,
                         show.legend = FALSE) +
  theme_minimal() +
  scale_fill_brewer(palette = "Spectral") +
  scale_color_brewer(palette = "Spectral") +
  theme_minimal(base_family = "Arial") +
  theme(legend.position = "top",
        legend.justification = "left",
        legend.title = element_text(face = "italic"),
        plot.background = element_rect(fill = "white", color = "grey88"),
        axis.text = element_text(color = "grey10"),
        title = element_text(face = "bold"),
        panel.grid = element_line(color = "grey94"),
        axis.title = element_text(hjust = 1),
        plot.subtitle = element_text(face = "italic"),
        plot.caption = element_text(face = "italic", size = 8),
        strip.text = element_text(hjust = 0, face = "bold")) +
  labs(title = "The most slowly adopted acts in EU history",
       subtitle = "Number of days between first Commission proposal and adoption",
       caption = "Hover over the points to display the title",
       x = "Date of adoption",
       y = "Days to adoption")

girafe(ggobj = iplot_outliers,
       fonts = list(sans = "Arial"),
       width_svg = 11,
       height_svg = 8,
       options = list(opts_sizing(rescale = FALSE),
                      opts_toolbar(saveaspng = FALSE),
                      opts_tooltip(css = "background-color:gray;color:white;font-style:italic;padding:10px;border-radius:5px;",
                                   use_fill = TRUE),
                      opts_hover_inv(css = "opacity:0.1;"),
                      opts_hover(css = "fill:green;"))
)
  
```

It turns out that many of the slowest legislative processes concerned – and this is particularly visible recently – the conclusion of international agreements. In these situations, the speed does not depend solely on the EU, as an agreement can only be formally concluded once the legal text is successfully negotiated with a third party.

Most acts on the list of slowest legislative files took well over ten years to complete.

### Regression {.tabset}

Regression analysis unpacks the relationship between a response (dependent) variable and one or more explanatory variables. In our case, we are interested in the relationship (or absence thereof) between the number of days it takes the EU to turn a Commission proposal into law and the number of Member States.

We will assume the relationship between the variables is functionally linear.

#### Table

```{r regression}

# number of member states over time
ms_years <- bind_rows(
  data.frame(
    year = 1952:1972,
    n_ms = 6
  ),
  data.frame(
    year = 1973:1980,
    n_ms = 9
  ),
  data.frame(
    year = 1981:1985,
    n_ms = 10
  ),
  data.frame(
    year = 1986:1994,
    n_ms = 12
  ),
  data.frame(
    year = 1995:2004,
    n_ms = 15
  ),
  data.frame(
    year = 2005:2006,
    n_ms = 25
  ),
  data.frame(
    year = 2007:2013,
    n_ms = 27
  ),
  data.frame(
    year = 2014:2018,
    n_ms = 28
  ),
  data.frame(
    year = 2019:2025,
    n_ms = 27
  )
)

# append N member states
days_ms <- acts_days |> 
  left_join(ms_years, by = c("year_adopted"="year")) |> 
  mutate(type = relevel(as.factor(type), ref = "Other"))

# regress
models <- list()

## m0
models[['Baseline']] <- lm(
  data = days_ms,
  formula = days ~ n_ms
)
names(models[['Baseline']]$coefficients)[1] <- "Constant"

## m1
models[['Controls']] <- lm(
  data = days_ms,
  formula = days ~ n_ms + type
)
names(models[['Controls']]$coefficients)[1] <- "Act type = Other"

## m2
models[['Zero Intercept']] <- lm(
  data = days_ms,
  formula = days ~ n_ms + type + 0
)

# effect size
effect <- as.numeric(round(models[['Controls']]$coefficients[which(names(models[['Controls']]$coefficients) == "n_ms")],2))

# summary for p values
mod_sum <- summary(models[['Controls']])

# p value
p_value <- mod_sum$coefficient[names(mod_sum$coefficient[,1]) == "n_ms", 4]

# p threshold
p_thresh <- 0.05

# can we reject the null
reject_null <- ifelse(effect > 0 & p_value < p_thresh, "can reject", "cannot reject")

# show table
modelsummary(models, 
             stars = TRUE,
             coef_map = c("n_ms" = "Number of Member States",
                          "Constant" = "Constant",
                          "typeOther" = "Act type = Other",
                          "Act type = Other" = "Act type = Other",
                          "typeRegulation" = "Act type = Regulation",
                          "typeDecision" = "Act type = Decision",
                          "typeDirective" = "Act type = Directive"),
             title = "Linear regression model of legislative efficiency (Y_i = days)",
             notes = list("In the 'Controls' specification, 'Other' is the reference category."))

```

#### Plot
```{r regviz}
modelplot(models[['Controls']])
```

#### Code

```{r regression, echo=TRUE, eval=FALSE}
```

#### Dataframe
```{r regdf}
days_ms
```

### {-}

Controlling for the type of act, the linear model predicts that for every additional Member State, the EU takes ```r effect``` days longer to adopt the legal act.

This model is far too simple to establish the existence of a causal relationship between the number of Member States and legislative efficiency. Indeed, some scholars would argue that no amount of conditioning on observables is sufficient to establish causality.

Nonetheless, our simple analysis suggests that we ```r reject_null``` the null hypothesis that more Member States in the EU system are not associated with slower law-making. 
