---
title: "Legislative Output of the European Union, Live!"
author: Michal Ovádek
output: 
  html_document:
    toc: true
    toc_float: true
    theme: yeti
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      error = FALSE)
```


```{r, include=FALSE}
library(eurlex)
library(ggplot2)
library(dplyr)
library(stringr)
library(rmarkdown)

acts <- elx_make_query("any", sector = 3) |> elx_run_query()

regs <- acts |> filter(str_sub(celex,6,6) == "R")
decs <- acts |> filter(str_sub(celex,6,6) == "D")
dirs <- acts |> filter(str_sub(celex,6,6) == "L")
recs <- acts |> filter(str_sub(celex,6,6) == "H")
```

The European Union (EU) is often criticized for a heavy regulatory touch. What is the reality? This automatically updated research document gives a real-time overview of the EU's legislative output.

Up to ```r format(Sys.time(), '%d %B, %Y')```, the EU and its predecessors have produced ```r nrow(regs)``` number of regulations, ```r nrow(dirs)``` number of directives, ```r nrow(decs)``` number of decisions and ```r nrow(recs)``` number of recommendations, according to [Eur-Lex](https://eur-lex.europa.eu/homepage.html) data pulled via the [eurlex](https://github.com/michalovadek/eurlex) package for R and a SPARQL API maintained by the Publications Office of the European Union.

Here is how the number of the four main legal acts changed over the lifespan of the EU.

```{r counttime}

n_relevant <- bind_rows(Regulations = regs, Decisons = decs, Directives = dirs, Recommendations = recs,
                        .id = "type") |> 
  mutate(year = as.integer(str_sub(celex, 2,5))) |> 
  count(year, type)

ggplot(n_relevant, aes(x = year, y = n, fill = type, color = type)) +
  geom_col(show.legend = FALSE) +
  scale_fill_brewer(palette = "Spectral") +
  scale_color_brewer(palette = "Spectral") +
  facet_wrap(~type, scale = "free_y") +
  theme_minimal() +
  theme(legend.position = "top",
        legend.justification = "left",
        legend.title = element_text(face = "italic"),
        plot.background = element_rect(fill = "grey30"),
        panel.grid = element_line(color = "grey28"),
        axis.text = element_text(color = "grey10"),
        title = element_text(face = "bold"),
        strip.text = element_text(hjust = 0, face = "bold")) +
  labs(x = NULL,
       y = "Number of acts",
       title = "Number of legal acts produced by the European Union",
       subtitle = "Source: Cellar API")

```

Proportion of each act over time

So we see there is significant over-time variation in the use of the different types of legal acts, but some types of acts, notably regulations, have been historically significantly more common than other types of acts. We can visualize the evolution of proportions directly:

```{r proptime}

expand.grid(year = min(n_relevant$year):max(n_relevant$year),
            type = unique(n_relevant$type)) |> 
  left_join(n_relevant) |> 
  mutate(n = ifelse(is.na(n), 0L, n)) |> 
  group_by(year) |> 
  mutate(total_y = sum(n),
         proportion = n/total_y) |> 
  ungroup() |> 
  ggplot(aes(x = year, y = proportion, fill = type, color = type)) +
  geom_area(show.legend = TRUE) +
  scale_fill_brewer(palette = "Spectral") +
  scale_color_brewer(palette = "Spectral") +
  theme_minimal() +
  theme(legend.position = "top",
        legend.justification = "left",
        legend.title = element_text(face = "italic"),
        plot.background = element_rect(fill = "grey30"),
        panel.grid = element_blank(),
        axis.text = element_text(color = "grey10"),
        title = element_text(face = "bold"),
        strip.text = element_text(hjust = 0, face = "bold")) +
  scale_x_continuous(expand = c(0,0)) +
  labs(x = NULL,
       y = NULL,
       title = "Proportion of each type of legal act by year")

```


